### 2024-04-05 appendices/app_EnergyScope.tex
RobPol - Description case ROB
Intro post-review AR

The objective, Eq.~(\ref{eq:obj_func_app}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Eq.~(\ref{eq:tau_app}).

@@

The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Eq.~(\ref{eq:c_inv_app}).  The installed capacity is defined with respect to the main end-uses output type, such as electricity for \gls{PV} or heat for a boiler. The total operation and maintenance costs (\textbf{C\textsubscript{maint}}) are calculated in the same way, Eq.~(\ref{eq:c_maint_app}).

@@

The total cost of the resources (\textbf{C\textsubscript{op}}) is calculated as the sum of the end-use over the different time-periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Eq.~(\ref{eq:c_op_app}). To simplify the reading, we write the sum over typical days as $t\in T$ such as in Eq.~(\ref{eq:c_op_app}). The period $T$ represents the sequence of hours and typical days over a year (8760h)\footnote{The exception is storage level which is optimised over the 365 days of the year instead of typical days.}. The full formulation is detailed in \cite{limpens2019energyscope} or in the documentation \cite{readthedocs_estd_v2.2}.

@@

The global annual \gls{GHG} emissions are calculated using a \gls{LCA} approach, i.e. taking into account emissions of the resources `\emph{from cradle to use}'. It is based on the indicator `\emph{GWP100a-IPCC2013}´ developed by the \acrfull{IPCC} \cite{IPCC2014}. For climate change, the natural choice as indicator is the \acrlong{GWP}, expressed in ktCO$_2$-eq./year. In Eq.~(\ref{eq:GWP_tot_app}), the total yearly emissions  of the system (\textbf{GWP\textsubscript{tot}}) are defined as the emissions related to resources (\textbf{GWP\textsubscript{op}}). The total emissions of the resources are the emissions associated to fuels (from cradle to combustion) and imports of electricity (\emph{gwp\textsubscript{op}}) multiplied by the period duration  (\emph{t\textsubscript{op}}), Eq.~(\ref{eq:GWP_op_app}).
Thus, this version accounts only for operation without accounting for the \gls{GWP} emitted during the construction of the technologies. This makes the results comparable with metrics used in the reports by the European Commission and the \acrfull{IEA}.

The above equations (Eqs. (\ref{eq:obj_func_app}) - (\ref{eq:GWP_op_app})) represent only a part of the formulation and illustrate the syntax that is used. Those representing the energy balance, network implementation, sectors representation, etc. are not presented in this work but are detailed in the latest version of the model, see \cite{limpens2021generating} and in the documentation \cite{readthedocs_estd_v2.2}.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{path_e.g._tech.png}
\caption{Example of how the technology capacities and associated variables are evolving. The example uses a technology with a 20 years lifetime. Initially 1~GW of capacity exists (\textbf{F\textsubscript{new}} during phase $2015\_2020$). Then another 1~GW is deployed (\textbf{F\textsubscript{new}} during phase $2020\_2025$). 15 years later, a part of the capacity reaches its lifetime limit and is removed (\textbf{F\textsubscript{old}} phase $2035\_2040$). 
Moreover, during the latter phase, additional capacity is decommissioned  prematurely (\textbf{F\textsubscript{decom}}). Finally, the technology reaches its expected lifetime and is fully withdrawn (\textbf{F\textsubscript{old}}).}
\label{fig:path_eg_igcc}
\end{figure}

@@

Similarly to a mass balance, Eq.~(\ref{eq:F_newBuilt_app}) is the technology capacity balance. The constraint forces the installation or withdrawing of capacities between two representative years: 
at the end of the phase (\emph{y\textsubscript{stop}}), the available capacity is the one used in the next representative year (\textbf{F}(\emph{y\textsubscript{stop}})). This capacity is equal to the one available in the previous representative year (\textbf{F}(\emph{y\textsubscript{start}})) plus the new installed capacity (\textbf{F\textsubscript{new}}) minus the capacity that has reached its lifetime (\textbf{F\textsubscript{old}}) minus the early decommissioned capacity (\textbf{F\textsubscript{decom}}). One notices that the capacity available for each representative year depends on a year (\emph{y\textsubscript{start}} or \emph{y\textsubscript{stop}}), while the other capacity changes depend on a phase (\emph{p} or \emph{p}2). Moreover, the decommissioning term depends on another phase, which is the one when the technology decommissioned has been built. As an illustration, Figure \ref{fig:path_eg_igcc} gives an example where 0.5 GW of a capacity built in 2015\_2020 is decommissioned in 2030\_2035 (\textbf{F\textsubscript{decom}}(2030\_2035,2015\_2020,i)).

@@

In linear programming, a solution might be mathematically correct, while not making sense in practice. As an example, a technology could be decommissioned before being built ($ \emph{p} < \emph{p\textsubscript{built}}$). Eqs.~(\ref{eq:F_decomNonPhysic1_app}-\ref{eq:Fold_def_app})  allow preventing these non-sense while keeping the formulation linear. 
Eq.~(\ref{eq:F_decomNonPhysic1_app}) forces the decommissioned capacity to zero when technology will be built after.
To do so, a parameter (\emph{decom\textsubscript{allowed}}) is defined \emph{a priori} and is equal to 0 or 1 when decommissioning is not possible or possible, respectively. 
Eq.~(\ref{eq:Fold_def_app}) defines the capacity reaching its lifetime limit at a certain phase, the concept is illustrated in Figure \ref{fig:path_eg_igcc}. For each phase, a set (\emph{AGE}) is calculated \emph{a priori}. It relates, for a given phase and technology, when the technology was built.  
In the case the technology has already reached its lifetime limit, the set (\emph{AGE}) returns the phase when the technology has been built. The first part of Eq.~(\ref{eq:Fold_def_app}) indicates that the technology is still available, and thus no capacity needs to be removed. The second part of the equation represents the capacity that reached its expected lifetime minus a part of the capacity that would have been decommissioned. 
As an example, Figure \ref{fig:path_eg_igcc} shows a 20 years lifetime technology with 1~GW of capacity installed before 2020. 
The `if' in Eq. (\ref{eq:Fold_def_app}) is linear as it is applied to a parameter and not a variable.

@@

To initialise the problem in 2020 with the existing design, an additional phase `\emph{2015\_2020}' is created. Eq.~(\ref{eq:Phase2020Design_app}) requires that the capacity used in 2020 is installed in the previous phase. \\

@@

Eq.~(\ref{eq:delta_tech_during_phase_app}) calculates the upper limit of change (\textbf{$\Delta$\textsubscript{change}}) in terms of supplied demand instead of installed capacity. 
%% GL_correction_v2 %% Following line added:
%  Ici on prend la différence entre ystart et ystop (et non pas le contraire), en effet, on est intéressé par la réticence des personnes à se séparer de leur ancienne technologie, plutot que la réticense à accepter une nouvelle techno. 
%% GL_correction_v2 %% end
Based on this quantification, the amount of change per phase is limited for low temperature heat (\emph{lim\textsubscript{LT,ren}}), Eq.~(\ref{eq:limit_reno_LTheat_app}), passenger mobility (\emph{lim\textsubscript{MobPass}}), Eq.~(\ref{eq:limit_reno_passmob_app}) and freight mobility (\emph{lim\textsubscript{MobFreight}}), Eq.~(\ref{eq:limit_reno_freight_app}). 
For instance, if the maximum allowable variation in supplied low temperature heat is set at 25\%, it would restrict the technology-related changes in low temperature heat to 25\% within a given phase. Consequently, if a technology supplies more than 25\% of the low temperature heat, it would require multiple phases to replace it with a different technology.\\
%As an example, setting the maximum change of supplied heat at low temperature to 25\% would limit the change in technology related to heat at low temperature to 25\% during a phase. 
%In this case, if more than 25\% of the low temperature heat was supplied by a technology, it would take more than one phase to replace it by a different one.

@@

As an extension of Eq. \ref{eq:obj_func_app}, the objective function to be minimised is the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}), 
defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Eq.~(\ref{eq:obj_func_v2_app}). The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Eq.~(\ref{eq:Capex_v2_app}), to which the residual asset value in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}). Thus, the investments account for the installation and dismantlement costs of the technologies. The total \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}) is the sum of the \gls{OPEX} in 2020 and the annualised sum of the \gls{OPEX} during each phase (\textbf{C\textsubscript{opex}}), Eq.~(\ref{eq:Copex_tot_v2_app}). During a phase, the system \gls{OPEX} is 
the product of the annualised phase factor, defined in Eq.~(\ref{eq:path_annu_factor_app}), and the arithmetic average of \gls{OPEX} cost for the representative years before and after the phase. The annualised phase factor is defined based on an average interest rate during the transition.
%assumed to be the annualised average of its first and last year. The annualised phase factor  ($\tau$\textsubscript{\emph{phase}}) is defined in Eq.~(\ref{eq:path_annu_factor_app}).

@@

For each year, the yearly \gls{OPEX} (\textbf{C\textsubscript{opex}}) is the sum of the operating and maintenance costs of technologies (\textbf{C\textsubscript{maint}}) and  the operating cost of the resources (\textbf{C\textsubscript{op}}), Eq.~(\ref{eq:opex_yearly_app}).

@@

The investment during a phase (\textbf{C\textsubscript{inv,phase}}) results from the multiplication of the newly built technologies (\textbf{F\textsubscript{new}}) with their annualised arithmetic averaged specific cost, Eq.~(\ref{eq:PhaseInv_app}). The annualised phase factor  (defined by Eq.~(\ref{eq:path_annu_factor_app})) is used. The specific cost during the phase is defined as  the average between the investment cost for the first and last year of the period.

@@

A part of the investment will remain after 2050. This residual investment, also called salvage value, can be calculated for each technology. 
A parameter, calculated \textit{a priori} , gives for each technology and construction phase, the remaining amount of years (\emph{remaining\_years}). 
As an example, if a PV panel has been built in 2045 and has a 20 years lifetime, the parameter will equal to 15 years. 
Thus, the salvage value is a fraction of the investment cost of this technology when it has been built. 
This fraction is the ratio between the number of remaining years and the lifetime of the technology. 
In the previous example, the residual investment of the PV built is 75\%. 
Eq. (\ref{eq:salvage_app}) computes, for each technology, the residual value that must be subtracted from the total cost. The residual value reflects the fact that the technology can still be used after the horizon of the model and is not fully amortised. The residual value is not applied to technologies that are removed prematurely. This differ from other models, such as Plexos where a technology removed prematurely will benefit from its salvage value (see analysis of \cite{waucquez2023validation}).

@@

The total \acrfull{GWP} emissions during the transition (\textbf{GWP\textsubscript{tot,trans}}) are equal to the sum of the total emissions per period (\textbf{GWP\textsubscript{tot}}), Eq.~(\ref{eq:gwp_tot_transition_app}). The emissions during a phase is estimated as the arithmetic average of the representative years before and after the phase. Eq.~(\ref{eq:limit_gwp_trans_app}) limits the total \gls{GWP} emissions during the transition by a maximum (\emph{gwp\textsubscript{lim,trans}}).

>>>>

The objective, Equation (\ref{eq:obj_func_app}), is the minimisation of the total annual cost of the energy system (\textbf{C\textsubscript{tot}}), defined as the sum of the annualised investment cost of the technologies (\textbf{$\tau$}$\cdot$\textbf{C\textsubscript{inv}}), the operating and maintenance costs of the technologies (\textbf{C\textsubscript{maint}}) and the operating cost of the resources (\textbf{C\textsubscript{op}}).
The annualised factor $\tau$ is computed \emph{a priori} based on the interest rate (\emph{i\textsubscript{rate}}) and the technology lifetime, (\emph{lifetime}), Equation (\ref{eq:tau_app}).

@@

The total investment cost (\textbf{C\textsubscript{inv}}) of each technology results from the multiplication of its specific investment cost (\emph{c\textsubscript{inv}}) and its installed capacity (\textbf{F}), see Equation (\ref{eq:c_inv_app}).  The installed capacity is defined with respect to the main end-uses output type, such as electricity for \gls{PV} or heat for a boiler. The total operation and maintenance costs (\textbf{C\textsubscript{maint}}) are calculated in the same way, Equation (\ref{eq:c_maint_app}).

@@

The total cost of the resources (\textbf{C\textsubscript{op}}) is calculated as the sum of the end-use over the different time-periods multiplied by the period duration (\emph{t\textsubscript{op}}) and the specific cost of the resources (\emph{c\textsubscript{op}}), Equation (\ref{eq:c_op_app}). To simplify the reading, we write the sum over typical days as $t\in T$ such as in Equation (\ref{eq:c_op_app}). The period $T$ represents the sequence of hours and typical days over a year (8760h)\footnote{The exception is storage level which is optimised over the 365 days of the year instead of typical days.}. The full formulation is detailed in \cite{limpens2019energyscope} or in the documentation \cite{readthedocs_estd_v2.2}.

@@

The global annual \gls{GHG} emissions are calculated using a \gls{LCA} approach, i.e. taking into account emissions of the resources `\emph{from cradle to use}'. It is based on the indicator `\emph{GWP100a-IPCC2013}´ developed by the \acrfull{IPCC} \cite{IPCC2014}. For climate change, the natural choice as indicator is the \acrlong{GWP}, expressed in ktCO$_2$-eq./year. In Equation (\ref{eq:GWP_tot_app}), the total yearly emissions  of the system (\textbf{GWP\textsubscript{tot}}) are defined as the emissions related to resources (\textbf{GWP\textsubscript{op}}). The total emissions of the resources are the emissions associated to fuels (from cradle to combustion) and imports of electricity (\emph{gwp\textsubscript{op}}) multiplied by the period duration  (\emph{t\textsubscript{op}}), Equation (\ref{eq:GWP_op_app}).
Thus, this version accounts only for operation without accounting for the \gls{GWP} emitted during the construction of the technologies. This makes the results comparable with metrics used in the reports by the European Commission and the \acrfull{IEA}.

The above equations (Equations (\ref{eq:obj_func_app}) - (\ref{eq:GWP_op_app})) represent only a part of the formulation and illustrate the syntax that is used. Those representing the energy balance, network implementation, sectors representation, etc. are not presented in this work but are detailed in the latest version of the model, see \cite{limpens2021generating} and in the documentation \cite{readthedocs_estd_v2.2}.

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=\textwidth]{path_e.g._tech.pdf}
\caption{Example of how the technology capacities and associated variables are evolving. The example uses a technology with a 20 years lifetime. Initially 1~GW of capacity exists (\textbf{F\textsubscript{new}} during phase $2015\_2020$). Then another 1~GW is deployed (\textbf{F\textsubscript{new}} during phase $2020\_2025$). 15 years later, a part of the capacity reaches its lifetime limit and is removed (\textbf{F\textsubscript{old}} phase $2035\_2040$). 
Moreover, during the latter phase, additional capacity is decommissioned  prematurely (\textbf{F\textsubscript{decom}}). Finally, the technology reaches its expected lifetime and is fully withdrawn (\textbf{F\textsubscript{old}}).}
\label{fig:path_eg_igcc}
\end{figure}

@@

Similarly to a mass balance, Equation (\ref{eq:F_newBuilt_app}) is the technology capacity balance. The constraint forces the installation or withdrawing of capacities between two representative years: 
at the end of the phase (\emph{y\textsubscript{stop}}), the available capacity is the one used in the next representative year (\textbf{F}(\emph{y\textsubscript{stop}})). This capacity is equal to the one available in the previous representative year (\textbf{F}(\emph{y\textsubscript{start}})) plus the new installed capacity (\textbf{F\textsubscript{new}}) minus the capacity that has reached its lifetime (\textbf{F\textsubscript{old}}) minus the early decommissioned capacity (\textbf{F\textsubscript{decom}}). One notices that the capacity available for each representative year depends on a year (\emph{y\textsubscript{start}} or \emph{y\textsubscript{stop}}), while the other capacity changes depend on a phase (\emph{p} or \emph{p}2). Moreover, the decommissioning term depends on another phase, which is the one when the technology decommissioned has been built. As an illustration, Figure \ref{fig:path_eg_igcc} gives an example where 0.5 GW of a capacity built in 2015\_2020 is decommissioned in 2030\_2035 (\textbf{F\textsubscript{decom}}(2030\_2035,2015\_2020,i)).

@@

In linear programming, a solution might be mathematically correct, while not making sense in practice. As an example, a technology could be decommissioned before being built ($ \emph{p} < \emph{p\textsubscript{built}}$). Equations (\ref{eq:F_decomNonPhysic1_app}-\ref{eq:Fold_def_app})  allow preventing these non-sense while keeping the formulation linear. 
Equation (\ref{eq:F_decomNonPhysic1_app}) forces the decommissioned capacity to zero when technology will be built after.
To do so, a parameter (\emph{decom\textsubscript{allowed}}) is defined \emph{a priori} and is equal to 0 or 1 when decommissioning is not possible or possible, respectively. 
Equation (\ref{eq:Fold_def_app}) defines the capacity reaching its lifetime limit at a certain phase, the concept is illustrated in Figure \ref{fig:path_eg_igcc}. For each phase, a set (\emph{AGE}) is calculated \emph{a priori}. It relates, for a given phase and technology, when the technology was built.  
In the case the technology has already reached its lifetime limit, the set (\emph{AGE}) returns the phase when the technology has been built. The first part of Equation (\ref{eq:Fold_def_app}) indicates that the technology is still available, and thus no capacity needs to be removed. The second part of the equation represents the capacity that reached its expected lifetime minus a part of the capacity that would have been decommissioned. 
As an example, Figure \ref{fig:path_eg_igcc} shows a 20 years lifetime technology with 1~GW of capacity installed before 2020. 
The `if' in Equation (\ref{eq:Fold_def_app}) is linear as it is applied to a parameter and not a variable.

@@

To initialise the problem in 2020 with the existing design, an additional phase `\emph{2015\_2020}' is created. Equation (\ref{eq:Phase2020Design_app}) requires that the capacity used in 2020 is installed in the previous phase. \\

@@

Equation (\ref{eq:delta_tech_during_phase_app}) calculates the upper limit of change (\textbf{$\Delta$\textsubscript{change}}) in terms of supplied demand instead of installed capacity. 
%% GL_correction_v2 %% Following line added:
%  Ici on prend la différence entre ystart et ystop (et non pas le contraire), en effet, on est intéressé par la réticence des personnes à se séparer de leur ancienne technologie, plutot que la réticense à accepter une nouvelle techno. 
%% GL_correction_v2 %% end
Based on this quantification, the amount of change per phase is limited for low temperature heat (\emph{lim\textsubscript{LT,ren}}), Equation (\ref{eq:limit_reno_LTheat_app}), passenger mobility (\emph{lim\textsubscript{MobPass}}), Equation (\ref{eq:limit_reno_passmob_app}) and freight mobility (\emph{lim\textsubscript{MobFreight}}), Equation (\ref{eq:limit_reno_freight_app}). 
For instance, if the maximum allowable variation in supplied low temperature heat is set at 25\%, it would restrict the technology-related changes in low temperature heat to 25\% within a given phase. Consequently, if a technology supplies more than 25\% of the low temperature heat, it would require multiple phases to replace it with a different technology.\\
%As an example, setting the maximum change of supplied heat at low temperature to 25\% would limit the change in technology related to heat at low temperature to 25\% during a phase. 
%In this case, if more than 25\% of the low temperature heat was supplied by a technology, it would take more than one phase to replace it by a different one.

@@

As an extension of Equation \ref{eq:obj_func_app}, the objective function to be minimised is the total transition cost of the energy system (\textbf{C\textsubscript{tot,trans}}), 
defined as the sum of the total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) and the \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}), according to Equation (\ref{eq:obj_func_v2_app}). The total \gls{CAPEX} (\textbf{C\textsubscript{tot,capex}}) is the sum of the investment during each phase (\textbf{C\textsubscript{inv,phase}}), Equation (\ref{eq:Capex_v2_app}), to which the residual asset value in 2050 is withdrawn (\textbf{C\textsubscript{inv,return}}). Thus, the investments account for the installation and dismantlement costs of the technologies. The total \gls{OPEX} (\textbf{C\textsubscript{tot,opex}}) is the sum of the \gls{OPEX} in 2020 and the annualised sum of the \gls{OPEX} during each phase (\textbf{C\textsubscript{opex}}), Equation (\ref{eq:Copex_tot_v2_app}). During a phase, the system \gls{OPEX} is 
the product of the annualised phase factor, defined in Equation (\ref{eq:path_annu_factor_app}), and the arithmetic average of \gls{OPEX} cost for the representative years before and after the phase. The annualised phase factor is defined based on an average interest rate during the transition.
%assumed to be the annualised average of its first and last year. The annualised phase factor  ($\tau$\textsubscript{\emph{phase}}) is defined in Eq.~(\ref{eq:path_annu_factor_app}).

@@

For each year, the yearly \gls{OPEX} (\textbf{C\textsubscript{opex}}) is the sum of the operating and maintenance costs of technologies (\textbf{C\textsubscript{maint}}) and  the operating cost of the resources (\textbf{C\textsubscript{op}}), Equation (\ref{eq:opex_yearly_app}).

@@

The investment during a phase (\textbf{C\textsubscript{inv,phase}}) results from the multiplication of the newly built technologies (\textbf{F\textsubscript{new}}) with their annualised arithmetic averaged specific cost, Equation (\ref{eq:PhaseInv_app}). The annualised phase factor  (defined by Equation (\ref{eq:path_annu_factor_app})) is used. The specific cost during the phase is defined as  the average between the investment cost for the first and last year of the period.

@@

A part of the investment will remain after 2050. This residual investment, also called salvage value, can be calculated for each technology. 
A parameter, calculated \textit{a priori} , gives for each technology and construction phase, the remaining amount of years (\emph{remaining\_years}). 
As an example, if a PV panel has been built in 2045 and has a 20 years lifetime, the parameter will equal to 15 years. 
Thus, the salvage value is a fraction of the investment cost of this technology when it has been built. 
This fraction is the ratio between the number of remaining years and the lifetime of the technology. 
In the previous example, the residual investment of the PV built is 75\%. 
Equation (\ref{eq:salvage_app}) computes, for each technology, the residual value that must be subtracted from the total cost. The residual value reflects the fact that the technology can still be used after the horizon of the model and is not fully amortised. The residual value is not applied to technologies that are removed prematurely. This differ from other models, such as Plexos where a technology removed prematurely will benefit from its salvage value (see analysis of \cite{waucquez2023validation}).

@@

The total \acrfull{GWP} emissions during the transition (\textbf{GWP\textsubscript{tot,trans}}) are equal to the sum of the total emissions per period (\textbf{GWP\textsubscript{tot}}), Equation (\ref{eq:gwp_tot_transition_app}). The emissions during a phase is estimated as the arithmetic average of the representative years before and after the phase. Equation (\ref{eq:limit_gwp_trans_app}) limits the total \gls{GWP} emissions during the transition by a maximum (\emph{gwp\textsubscript{lim,trans}}).

### 2024-04-05 appendices/app_case_study.tex
Intro post-review AR

The defossilisation of the different sectors are not performed at the same rate. The non-energy demand of methanol and ammonia are substituted by electrofuels. These are the first use of electrofuels as  e-ammonia is the cheapest electrofuel thanks to the high maturity of the Haber-Bosch process. The decentralised heat and mobility sectors are also dropping first. This is a combination of efficiency and substitution of fossil fuels with electricity. In particular, the private mobility drops to near-zero emissions at early stages of the transition for two main reasons: (i) the switch of propulsion systems and, (ii) the modal shift. First, while accounting for the society inertia in this sector (see Eq. \ref{eq:limit_reno_passmob_app}), battery electric vehicles (BEV) substitute \gls{ICE} cars. This reduces the overall emissions of the private vehicles as electric motors are more efficient than \gls{ICE} and they are driven by a less-emitting resource (\ie electricity versus diesel and gasoline). On top of this, the model allows a modal shift from the private to the public mobility. From the current share of the passenger mobility provided by private vehicles (\ie 80\%), the model allows this share to go down to 50\%. Efficiency comes mainly from district heating networks and electrical heat pumps for the heat sector, and public mobility and electric cars for the mobility sector. From 2040 onward, the decreases are mainly due to the substitution of the remaining fossil fuels by electrofuels as illustrated in Figure \ref{fig:pestd_primary_energy}.

@@

A part of the investment will be recovered at the end of the transition based on the remaining lifespan of the technologies after 2050. Figure \ref{fig:pestd_inv_return} illustrates the salvage value by sectors, calculated according to Eq. (\ref{eq:salvage}). Out of the 114.4~b€\textsubscript{2015} of investments in the infrastructure (\ie mostly power grid and gas network), 55.9\% remain available after 2050, due to their long lifetime. On the contrary, private mobility has a lower salvage value due to a major drop within the first four years and an average lifetime below 10 years \cite{febiac2021datadigest}. 
% Consequently, out the 240.5~b€\textsubscript{2015}, including the initial investment of 75.7~b€\textsubscript{2015}, only 13.9\% would remain available after 2050.

@@

In general, due to the formulation of the salvage value (see Eq. \ref{eq:salvage}), the myopic approach is more techno-oriented as investing more in technologies is beneficial, especially at the early stages of the transition. Therefore, before converging to a similar energy mix in 2050, the myopic system relies more on local renewables (\eg solar and wind) than on importing renewable energy carriers (\eg e-ammonia, e-methanol, e-hydrogen or e-methane), see Figure \ref{fig:my_pestd_res_cat_diff}. In parallel, in the near term, the system relies on average slightly more on conventional/non-renewable sources, like observed in other studies \cite{keppo2010short,nyqvist2005limited,hedenus2006induced}.

>>>>

The defossilisation of the different sectors are not performed at the same rate. The non-energy demand of methanol and ammonia are substituted by electrofuels. These are the first use of electrofuels as  e-ammonia is the cheapest electrofuel thanks to the high maturity of the Haber-Bosch process. The decentralised heat and mobility sectors are also dropping first. This is a combination of efficiency and substitution of fossil fuels with electricity. In particular, the private mobility drops to near-zero emissions at early stages of the transition for two main reasons: (i) the switch of propulsion systems and, (ii) the modal shift. First, while accounting for the society inertia in this sector (see Equation (\ref{eq:limit_reno_passmob_app})), battery electric vehicles (BEV) substitute \gls{ICE} cars. This reduces the overall emissions of the private vehicles as electric motors are more efficient than \gls{ICE} and they are driven by a less-emitting resource (\ie electricity versus diesel and gasoline). On top of this, the model allows a modal shift from the private to the public mobility. From the current share of the passenger mobility provided by private vehicles (\ie 80\%), the model allows this share to go down to 50\%. Efficiency comes mainly from district heating networks and electrical heat pumps for the heat sector, and public mobility and electric cars for the mobility sector. From 2040 onward, the decreases are mainly due to the substitution of the remaining fossil fuels by electrofuels as illustrated in Figure \ref{fig:pestd_primary_energy}.

@@

A part of the investment will be recovered at the end of the transition based on the remaining lifespan of the technologies after 2050. Figure \ref{fig:pestd_inv_return} illustrates the salvage value by sectors, calculated according to Equation (\ref{eq:salvage}). Out of the 114.4~b€\textsubscript{2015} of investments in the infrastructure (\ie mostly power grid and gas network), 55.9\% remain available after 2050, due to their long lifetime. On the contrary, private mobility has a lower salvage value due to a major drop within the first four years and an average lifetime below 10 years \cite{febiac2021datadigest}. 
% Consequently, out the 240.5~b€\textsubscript{2015}, including the initial investment of 75.7~b€\textsubscript{2015}, only 13.9\% would remain available after 2050.

@@

In general, due to the formulation of the salvage value (see Equation (\ref{eq:salvage})), the myopic approach is more techno-oriented as investing more in technologies is beneficial, especially at the early stages of the transition. Therefore, before converging to a similar energy mix in 2050, the myopic system relies more on local renewables (\eg solar and wind) than on importing renewable energy carriers (\eg e-ammonia, e-methanol, e-hydrogen or e-methane), see Figure \ref{fig:my_pestd_res_cat_diff}. In parallel, in the near term, the system relies on average slightly more on conventional/non-renewable sources, like observed in other studies \cite{keppo2010short,nyqvist2005limited,hedenus2006induced}.

### 2024-04-05 chap_RL/chap_RL.tex
Intro post-review AR

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.7\textwidth]{Reward.pdf}
\caption{Reward, $R$, function scheme. Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded. If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues. Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost.}
\label{fig:Reward}
\end{figure}

@@

Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the training of the \gls{NN}. For this, numerous episodes are played through the myopic optimization of the transition pathway of Belgium. At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}). Then, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window. This sample will remain valid for the following time windows. In other words, there is only one sample draw per episode (see Figure \ref{fig:ranges_transition}).  Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward. This goes on until the end of the episode. For the new episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.

To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase, also called ``training'', of the agent has been done on the monthly model. Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus about 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.7\textwidth]{Binding_constr.pdf}
\caption{Binding versus non-binding constraints. In \gls{LP} where the feasibility domain is non-empty and bounded, the constraints defined a convex feasibility domain in the space of variables (here, x$_1$ and x$_2$). The optimal solution usually locates on a vertex of this domain, \ie the intersection of several constraints (here, constraints 2 and 3) limiting the solution. These constraints are considered as binding, \ie having a limiting impact on the optimal solution.}
\label{fig:Reward}
\end{figure}

>>>>

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.7\textwidth]{Reward.pdf}
\caption{Reward function, $R$. Before reaching 2050, the episode is prematurely ended and a negative reward is given if the optimisation is infeasible or if the \ce{CO2}-budget is exceeded. If the optimisation provides a solution and the \ce{CO2}-budget is not exceeded, the episode continues. Finally, if the episode goes until 2050, the reward is a weighted sum between the capped cumulative emissions and the total transition cost.}
\label{fig:Reward}
\end{figure}

@@

Before testing the optimal policy $\pi^*\left(\bm{a}_n | \bm{o}_n\right)$, the first step consists in assessing the learning of the \gls{NN}, also called ``training''. For this, numerous episodes are played through the myopic optimisation of the transition pathway of Belgium. At the beginning of each episode, the agent starts with the actual Belgian energy system of 2020 (see Appendix \ref{app:bel_2020}). Then, a sample of values, drawn for the uncertain parameters, affects the model for the 2020-2030 time window. This sample will remain valid for the following time windows. In other words, there is only one sample draw per episode (see Figure \ref{fig:ranges_transition}).  Then, the agent takes a set of actions, affecting the environment that feeds back the agent with a new state and a reward. This goes on until the end of the episode. For the new episode, similarly to the \gls{UQ} analysis (see Section \ref{sec:meth:UQ}), the new sample of uncertain parameters for the new episode is drawn following the quasi-random Sobol' sampling technique \cite{bratley2003implementing}.

To reduce the computational burden and maximise the exploration of the transition pathways, the learning phase of the agent has been done on the monthly model. Even though averaging time series of end-use demands and renewable productions brings some discrepancies (\ie faster emergence of local \gls{VRES} and smaller electrification of the system), the main advantage of the monthly approach is its computational time, \ie couple of seconds versus about 15 min for the hourly model (see Appendix \ref{app:mo_vs_td}).

The learning phase has been split by batch of 500 steps. For the first 100 steps of each batch, the \gls{RL}-model collects transitions before learning starts. This makes sure replay buffer is full enough for useful updates. At the end of each batch, the up-to-date policy, \ie the \gls{NN}, is saved. This way, we can assess the progress in the learning process and its convergence (see Figure \ref{fig:reward_success}).

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.49\textwidth]{Mean_reward.pdf}
\includegraphics[width=0.49\textwidth]{Success_rate.pdf}
\caption{Mean reward and success rate of the different learning batches. The stabilisation of the reward curve shows a convergence of the learning process for the agent's point of view. The success rate also confirms this progression and convergence.}
\label{fig:reward_success}
\end{figure}

@@

\begin{figure}[!htbp]
\centering
\includegraphics[width=0.7\textwidth]{Binding_constr.pdf}
\caption{Binding versus non-binding constraints. In \gls{LP} where the feasibility domain is non-empty and bounded, the constraints defined a convex feasibility domain in the space of variables (here, x$_1$ and x$_2$). The optimal solution usually locates on a vertex of this domain, \ie the intersection of several constraints (here, constraints 2 and 3) limiting the solution. These constraints are considered as binding, \ie having a limiting impact on the optimal solution.}
\label{fig:Binding_constr}
\end{figure}
